---
title: "BDA3 Chapter 3 Exercise 3"
author: "Brian Callander"
date: "2018-09-16"
tags: bda chapter 3, solutions, bayes, normal, t, behrens-fischer
always_allow_html: yes
output: 
  md_document:
    variant: markdown
    preserve_yaml: yes
---

Here's my solution to exercise 3, chapter 3, of [Gelman's](https://andrewgelman.com/) *Bayesian Data Analysis* (BDA), 3rd edition. There are [solutions](http://www.stat.columbia.edu/~gelman/book/solutions.pdf) to some of the exercises on the [book's webpage](http://www.stat.columbia.edu/~gelman/book/).

<!--more-->

<div style="display:none">
  $\DeclareMathOperator{\dbinomial}{Binomial}
   \DeclareMathOperator{\dbern}{Bernoulli}
   \DeclareMathOperator{\dpois}{Poisson}
   \DeclareMathOperator{\dnorm}{Normal}
   \DeclareMathOperator{\dt}{t}
   \DeclareMathOperator{\dcauchy}{Cauchy}
   \DeclareMathOperator{\dexponential}{Exp}
   \DeclareMathOperator{\dgamma}{Gamma}
   \DeclareMathOperator{\dinvgamma}{InvGamma}
   \DeclareMathOperator{\invlogit}{InvLogit}
   \DeclareMathOperator{\logit}{Logit}
   \DeclareMathOperator{\ddirichlet}{Dirichlet}
   \DeclareMathOperator{\dbeta}{Beta}$
</div>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  error = TRUE,
  dev = "svglite",
  fig.ext = ".svg" 
)

library(tidyverse)
library(scales)
library(kableExtra)

theme_set(theme_bw())

```

Suppose we have $n$ measurements $y \mid \mu, \sigma \sim \dnorm(\mu, \sigma)$, where the prior $p(\mu, \log \sigma) \propto 1$ is uniform. The calculations on page 66 show that the marginal posterior distribution of $\mu$ is $\mu \mid y \sim \dt(\bar y, s / n)$, where $s$ is the sample standard deviation.


```{r data}
control <- list(
  n = 32,
  mean = 1.013,
  sd = 0.24
)

treatment <- list(
  n = 36,
  mean = 1.173,
  sd = 0.2
)
```

```{r}
library(LaplacesDemon)

tibble(value = seq(0.95, 1.25, 0.001)) %>% 
  mutate(
    ctrl = dst(value, control$mean, control$sd / control$n, control$n - 1),
    trt = dst(value, treatment$mean, treatment$sd / treatment$n, treatment$n - 1)
  ) %>% 
  gather(cohort, density, ctrl, trt) %>% 
  ggplot() +
  aes(value, density, fill = cohort) +
  geom_area(colour = 'white') +
  labs(
    x = 'Value',
    y = 'Density',
    title = 'Marginal posterior mean distributions'
  )
```

```{r}
draws <- 10000

difference <- tibble(draw = 1:draws) %>% 
  mutate(
    ctrl = rst(n(), control$mean, control$sd / control$n, control$n - 1),
    trt = rst(n(), treatment$mean, treatment$sd / treatment$n, treatment$n - 1),
    difference = trt - ctrl
  ) 
  
difference %>% 
  ggplot() +
  aes(difference) +
  geom_histogram() +
  labs(
    x = 'Difference',
    y = 'Count',
    title = 'Difference in marginal posteriors of the means',
    subtitle = 'Treatment - Control'
  )
```

```{r}
difference$difference %>% 
  quantile(probs = c(0.05, 0.95)) 
```

